---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Shuer Li"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
# Set up
library(tidyverse)
library(agricolae)
library(here)
here()

NTL_raw <- read.csv(
  here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors = TRUE
  )

# Set data to data format
NTL_raw$sampledate <- as.Date(NTL_raw$sampledate, format = "%m/%d/%y")

#2
# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no relationship between lake temperature and depth in July (beta = 0).
Ha: Lake temperature changes with depth in July (beta != 0).


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_July <- NTL_raw %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#5
TemperaturebyDepth <-
  ggplot(NTL_July, aes(x = depth, 
                       y = temperature_C)) +
  geom_point() +
  geom_smooth(method = "lm",
              col = "blue") +
  ylim(0, 35)
print(TemperaturebyDepth) 

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure indicates that temperature decreases steadily with increasing depth across all lakes, showing a strong negative linear relationship. The points are distributed in a pattern consistent with a linear decline, supporting the use of a linear regression model for this data.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
TempbyDepth.regression <- lm(temperature_C ~ depth, data = NTL_July)
summary(TempbyDepth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The linear regression shows a highly significant negative relationship between temperature and depth (p < 0.05). The model explains about 73.87% of the variance in temperature (R-squared = 0.7387), indicating that depth is a strong predictor of temperature across all lakes. According to the regression coefficient, for every 1-meter increase in depth, lake temperature decreases by approximately 1.94621 °C.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
TempAIC <- lm(data = NTL_July, temperature_C ~ year4 + daynum + depth)
step(TempAIC)

#10
TempModel <- lm(data = NTL_July, temperature_C ~ year4 + daynum + depth)
summary(TempModel)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method selected a model including year4, daynum, and depth as the best set of explanatory variables to predict lake temperature. This multiple regression model explains approximately 74.12% of the observed variance (R-squared = 0.7412) in temperature during July. Compared to a simpler model using only depth as the explanatory variable, this represents a clear improvement in explanatory power, as adding year and daynum slightly increases the amount of variance explained and captures additional temporal trends in the data.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Run ANOVA test
Temp.anova <- aov(data = NTL_July, temperature_C ~ lakename)
summary(Temp.anova)

# Run linear model
Temp.regression <- lm(data = NTL_July, temperature_C ~ lakename)
summary(Temp.regression)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The ANOVA results indicate a statistically significant difference in mean July temperature among the nine lakes (F = 50, p < 0.05). Therefore, we reject the null hypothesis that all lakes have the same mean temperature. The corresponding linear model confirmed this finding (R-squared = 0.03953), showing that while lake identity explains only a small portion of total temperature variation, the differences among lakes are statistically meaningful.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
TemperaturebyDepth2 <-
  ggplot(NTL_July, aes(x = depth, 
                       y = temperature_C,
                       color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",
              se = FALSE) +
  ylim(0, 35)
print(TemperaturebyDepth2) 


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Temp.anova)
Temp.groups <- HSD.test(Temp.anova, "lakename", group = TRUE)
Temp.groups


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Based on the Tukey’s HSD groupings, Peter Lake (“c”) shares the same mean July temperature with Paul Lake (“c”) and Ward Lake (“bc”), as they belong to overlapping statistical groups, indicating no significant difference between the three. All other lakes belong to different statistical groups (a, b, d, or e), meaning their mean temperatures are significantly different from Peter Lake. No single lake has a mean temperature that is statistically distinct from all other lakes, as every group overlaps with at least one other.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just comparing Peter Lake and Paul Lake, we could use a two-sample t-test to test whether the mean July temperatures of the two lakes differ significantly. This test compares the means of two independent groups and determines if the difference between them is statistically significant.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18
Crampton_Ward <- NTL_July %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))
Temp.twosample <- t.test(temperature_C ~ lakename, data = Crampton_Ward)
Temp.twosample

```

>Answer: The two-sample t-test comparing July temperatures between Crampton Lake and Ward Lake yielded t(200.37) = 1.1181, p = 0.2649. Because p > 0.05, we fail to reject the null hypothesis and conclude that the mean July temperatures of the two lakes are not significantly different. This finding is consistent with the Tukey’s HSD results in part 16, where Crampton Lake (“ab”) and Ward Lake (“bc”) shared overlapping group letters—indicating statistically similar mean temperatures.
